---
title: "Genome Informatics"
author: "Laura Sun (PID: A17923552)"
format: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1: Proportion of G/G in a population

Download a csv from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39876388-40046389;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read csv file for MXL:

```{r}
mxl <- read.csv("wk9/373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

Let's look at a different population:

```{r}
gbr <- read.csv("wk9/373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

```{r}
table(gbr$Genotype..forward.strand.)
```

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr) * 100,2)
```

This variant that is associated with child asthma is more frequent in the GBR population than MKL population.

Let's dig further:

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("wk9/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

Sample size of each genotype:

```{r}
table(expr$geno)
```
Median expression levels of each genotype:

```{r}
library(dplyr)
expr %>% group_by(geno) %>% summarize(median=median(exp))
```

```{r}
library(ggplot2)
```

Let's make a boxplot:

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```

From this plot, we see that the relative expression levels are A|A > A|G > G|G, with a relatively big difference. SNP does effect the expression of ORMDL3.