---
title: "Class10: Halloween mini project"
author: "Laura Sun (PID: A17923552)"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter, let's do a mini project to help us figure out the best candy!

Our data come from the 538 website and is available as a CSV file:

```{r}
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

or

```{r}
candy |>
  nrow()
```
or call the tidyverse package!

```{r}
library(tidyverse)
candy%>%
  nrow()
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Hershey's Kisses, and the winpercent value is 55.37545.

```{r}
candy["Hershey's Kisses", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

With tidyverse!

```{r}
library(dplyr)
candy |> filter(rownames(candy)=="Kit Kat") |> select(winpercent)
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Quick overview of the dataset

```{r}
library(skimr)
skim(candy)
## skimr::skim(candy) to be easier
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale. The rest are on 0-1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

The candy does not contain chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) + geom_histogram(bins=20)
```


> Q9. Is the distribution of winpercent values symmetrical?

It's not symmetrical. The graph is right skewed.

```{r}
ggplot(candy, aes(winpercent)) + geom_density()

```


> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

This depends on checking mean or median for center. Mean is above 50%, median is lower than 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset.
# 2. Find their winpercent values
# 3. Calculate the mean of these values.

#4-6 Do the same for fruity candy.
# 7. Compare the mean winpercents of chocolate vs. fruity
# 8. Pick the higher one as winner.

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds,]$winpercent
choc.mean <- mean(choc.win)
choc.mean

fr.inds <- candy$fruity==1
fr.win <- candy[fr.inds,]$winpercent
fr.mean <- mean(fr.win)
fr.mean
```

This is harder to read:
```{r}
mean(candy[candy$chocolate==1,]$winpercent)
mean(candy[candy$fruity==1,]$winpercent)
```
Another way:

```{r}
candy |> filter(chocolate==1) |> select(winpercent)
```

```{r}
choc.mean>fr.mean
```
On average, chocolate candy (60.92153) is higher ranked than fruit candy (44.11974).


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fr.win)
```

The difference is statistically significant. We have a p-value of 2.871e-08, which is a lot smaller than 0.05. We have enough evidence to support that there is a significant difference between chocolate and fruity candy's winpercent.

> Q13. What are the five least liked candy types in this set?

```{r}
candy |> arrange(winpercent) |> head(5)
```
or baseR:

```{r}
position <- order(candy$winpercent)
head(candy[position,],5)

```


Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```

or

```{r}
candy |> arrange(winpercent) |> tail(5)
```

or

```{r}
candy |> arrange(-winpercent) |> head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy, aes(winpercent, rownames(candy))) + geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + aes(x=winpercent, y=reorder(rownames(candy), winpercent)) + geom_col()
```

Adding colors based on the type of candy!

```{r}
my_cols <- rep("black",nrow(candy)) # returns a vector of "black", if used in graph, all black bars.
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$fruity)] <- "pink"
my_cols[as.logical(candy$bar)] <- "brown" #for chocolate bar, overide chocolate color with brown.

ggplot(candy) + aes(x=winpercent, y=reorder(rownames(candy), winpercent)) + geom_col(fill=my_cols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocolate candy.

> Q18. What is the best ranked fruity candy?

Starbust is the best ranked fruity candy.

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent.

```{r}
library(ggrepel)
my_cols[as.logical(candy$fruity)] <- "red"

ggplot(candy) + aes(winpercent, pricepercent, label=rownames(candy)) + 
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps=7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Based on the graph, Reese's Miniatures seem to be highly popular and affordable.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent)
tail(candy[ord,c(11,12)], n=5)
```

The five most expensive candies are listed above. Among them, Nik L Nip is the least popular (with a winpercent of around 22)

## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Our main pca result figure:

```{r}
ggplot(pca$x) + aes(PC1, PC2, label=rownames(pca$x)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols)
```

We should also examine the "variable loadings" or contributions of the orignial variables to the new PCs.

```{r}
ggplot(pca$rotation) + aes(PC1, rownames(pca$rotation)) + geom_col()
```


```{r}
p <- ggplot(pca$x) + aes(PC1, PC2, label=rownames(pca$x)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols)

```

Interactive plots than can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF :(


```{r}
library(plotly)
ggplotly(p)
```

