---
title: "Class 12: RNASeq Analysis"
author: "Laura Sun (PID:A17923552)"
format: pdf
toc: true
---

## Background
Today we will analyze some RNASeq data from himes et al. on the effects of a common steroid (dexamethasone) on airway smooth muscle cells (ASM cells).

Our starting point is the "counts" data and "metadata" that contain the count values for each gene in their different experiments (i.e. cell lines with or without the drug).

## Data import

Load in console:
install.packages("BiocManager")
BiocManager::install("DESeq2")
n

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

Let's take a look:

```{r}
head(counts)
```

> Q1: How many genes are in this dataset?

Number of genes:

```{r}
nrow(counts)
```

Experiments:

```{r}
nrow(metadata)
```


> Q2: How many ‘control’ cell lines do we have?

```{r}
sum(metadata$dex=="control")
```

## Toy differential gene expression

To start our analysis, let's calculate the mean counts for all genes in the "control" experiments.

1. Extract all control columns from the `counts` object.
2. Calculate the mean for all rows (i.e. genes) of these "control" columns.

3-4. Do the same for "treated" group
5. Compare these `control.mean` and `treated.mean` values.

```{r}
control.inds <- metadata$dex=="control"
control.counts <- counts[,control.inds]
control.means <- rowSums(control.counts)/4 # Getting the average of each gene's 4 controls. Or use rowMeans(control.counts)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

We can use `rowMeans()` instead of `rowSums()`/counts to make it easier.

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated.inds <- metadata$dex=="treated"
treated.counts <- counts[,treated.inds]
treated.means <- rowMeans(treated.counts)
```

Store these together for ease of bookkeeping as `meancounts`

```{r}
meancounts <- data.frame(control.means, treated.means)
head(meancounts)
```

Make a plot of control vs. treated mean values for all genes.

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?

Use geom_point() for scatter plot.

```{r}
library(ggplot2)
ggplot(meancounts, aes(x=control.means, y=treated.means)) + geom_point() + theme_bw()
```

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?

Use scale_x_log10() to adjust for skewed data.

```{r}
ggplot(meancounts, aes(x=control.means, y=treated.means)) + geom_point() + theme_bw() + scale_x_log10() + scale_y_log10()
# or plot(meancounts, log="xy")
```

We often talk about metrics like "log2 fold-change"

```{r}
# control/treated
log2(10/10) # gives 0, so 0 change.
```

```{r}
log2(40/10)
log2(10/40)
```

Let's calculate the log2 fold change for our treated oven control mean counts.

```{r}
meancounts$log2fc <- log2(meancounts$treated.means/ meancounts$control.means)
# add a column.

head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7: What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

The arr.ind=TRUE with which() return the indices of TRUE values, showing the location of zeros. We ignore genes with zero counts. unique() helps to not count a row twice if it has zeros.

A common "rule of thumb" is a log2 full change cutoff of +2 and -2 to call genes "Up regulated" or "Down regulated". 

> Q8: Number of "up regulated" genes:

```{r}
sum(meancounts$log2fc > +2, na.rm=T)
#na.rm=T means don't count the NAs.
```

> Q9: Number of down regulated genes:

```{r}
sum(meancounts$log2fc < -2, na.rm=T)
```

> Q10: We are missing statistical significance of the data. We don't know if the difference between control and treated are caused by random errors, skewed data, or there's a real significant difference. We can use something like two sample t-test to check for statistical significance.

## DESeq2 analysis

```{r, message=FALSE}
library(DESeq2)
```

For DESeq analysis we need three things:

1. count values (`countData`)
2. metadata telling us about the columns in `countData` (`colData`)
3. design of the experiment (i.e. what do you want to compare)

Our first function from DESeq2 will setup the input required for analysis by storing all these 3 things together.

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~dex)
```

The main function in DESeq2 that runs the analysis is called `DESeq()`

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

```{r}
36000*0.05
```

## Volcano Plot

This is a common summary result figure from these types of experiments and plot the log2 fold-change vs. the adjusted p-value.

```{r}
plot(res$log2FoldChange, -log(res$padj))
# use -log to change the axis

abline(v=c(-2,2), col="red")
abline(h=-log(0.04), col="red")
```


## Save our result

```{r}
write.csv(res, file="my_results.csv")
```

## Add gene annotation

To help make sense of our results and communicate them to other folk, we need to add more annotations to our main `res` object.

We will use two bioconductor packages to first map IDs to different formats including classic gene "symbol" gene name.

Install these with the following commands:
`BiocManager::install("AnnotationDbi")`
`BiocManager::install("org.Hs.eg.db")`

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
#mygenes$control <- maplds(org.Hs.eg.db, column="SYMBOL", #keys=row.names(mygenes), keytype"ENSEMBL")
```

Let's see what's in `org.Hs.eg.db` with the `columns()` function"

```{r}
columns(org.Hs.eg.db)
```

We can translate or "map" IDs between any of these 26 databases using the `mapIDs()` function.

```{r}
head(row.names(res))
res$symbol <- mapIds(keys=row.names(res), # our current IDs
       keytype="ENSEMBL", # the format of our IDs
       x=org.Hs.eg.db, # where to get the mappings from
       column="SYMBOL") # the format/DB to map to
head(res)
```

Add the mappings for "GENENAME" and "ENTREZID" and store as `res$genename` and `res$entrez`

```{r}
res$genename <- mapIds(keys=row.names(res), # our current IDs
       keytype="ENSEMBL", # the format of our IDs
       x=org.Hs.eg.db, # where to get the mappings from
       column="GENENAME") # the format/DB to map to

res$entrez <- mapIds(keys=row.names(res), # our current IDs
       keytype="ENSEMBL", # the format of our IDs
       x=org.Hs.eg.db, # where to get the mappings from
       column="ENTREZID") # the format/DB to map to
```

## Pathway Analysis

There are lots of bioconductor packages to do this type of analysis. For now let's try one called **gage**. Again, we need t o install this if we don't have it already.

BiocManager::install("gage")
BiocManager::install("gageData")
BiocManager::install("pathview")

```{r, message=FALSE}
library(gage)
library(gageData)
library(pathview)
```

To use **gage** I need two things

- a named vector of fold-change DEGs (our geneset of interest)
- a set of pathways of genesets to use for 

```{r}
x <- c("barry"=5, "lisa"=10)
x
```

```{r}
names(x) <- c("low", "high")
x
```

```{r}
foldchanges <- res$log2FoldChange
names(foldchanges) <- res$entrez
head(foldchanges)
```

```{r}
data(kegg.sets.hs)
keggres=gage(foldchanges,gsets=kegg.sets.hs)
```

In our results object we have:

```{r}
attributes(keggres)
```

```{r}
head(keggres$less,5)
```

Let's look at one of these pathways (hsa05310 Asthma) with our gene colored up so we can see the overlap

```{r}
pathview(pathway.id="hsa05310", gene.data=foldchanges)
```

Add this pathway figure to our lab report:

![]()

## Save our main results

```{r}
write.csv(res, file="myresults_annotated.csv")
```

