---
title: "Class07: Machine Learning 1"
author: "Laura Sun (A17923552)"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction.

## K-means clustering
To see how this works let's first makeup some data to cluster where we know what the answer should be. We can use the `rnorm()` function to help here:

```{r}
#(n, mean=0, sd=1)
hist(rnorm(500, mean=5))
```

```{r}
x <- c(rnorm(30, mean=-3), rnorm(30, mean=3))
y <- rev(x)
```

```{r}
x <- cbind(x,y)
plot(x)
```

The function for K-means clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x, centers=2)
k
```

To get the results of the returned list object we can use the dollar `$` syntax.

> Q. How many points are in each cluster?

```{r}
k$size
```

> Q. What 'component' of your result object details
  - cluster assignment/ membership?
  - cluster center?
  
```{r}
k$cluster
k$centers
```

> Make a clustering results figure of the data colored by cluster membership and show cluster centers.

```{r}
# plot(x, col=c("red", "blue")), not colored by cluster.
plot(x, col=k$cluster, pch=16)
points(k$centers, col="blue", pch=15, cex=2)
```

K-means clustering is very popular as it is very fast and relatively straight forward: it takes numeric data as input and returns the cluster membership vector etc.

The "issue" is we tell `kmeans()` how many clusters we want!

> Run kmeans again cluster into 4 groups/clusters and plot the results like we did above.

```{r}
k4 <- kmeans(x, centers=4)
plot(x, col=k4$cluster, pch=16)
points(k4$centers, col="blue", pch=15, cex=2)
k4
```

Scree plot to pick k `centers` value
Or brute force:

```{r}
k1 <- kmeans(x, centers=1)
k2 <- kmeans(x, centers=2)
k3 <- kmeans(x, centers=3)
k4 <- kmeans(x, centers=4)
k5 <- kmeans(x, centers=5)

z <- c(k1$tot.withinss,
       k2$tot.withinss,
       k3$tot.withinss,
       k4$tot.withinss,
       k5$tot.withinss)
plot(z, typ="b")
```


```{r}
n <- NULL
for(i in 1:5){
  n <- c(n,kmeans(x, centers=i)$tot.withinss)
}
plot(n, typ="b")
```

## Hierarchical Clustering

The main "base" R function for hierarchical clustering is called `hclust()`. here we can't just input our data we need to first calculate a distance matrix (e.g. `dist()`) for our data and use this as input to `hclust()`.

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

There is a plot method for hclust results. Let's try it:

```{r}
plot(hc)
abline(h=8, col="red")
```

1 to 30 on one side, 30 to 60 on the other side.
To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a give height or at a height that yields a given "k" group.

```{r}
cutree(hc, h=8)
```


```{r}
grps <- cutree(hc, k=2)
```

> Q. Plot the data with our hclust result coloring.

```{r}
plot(x, col=grps)
```

# Principal Component Analysis (PCA)

## PCA of UK Food Data

Import food data from an online CSV file:

```{r}
url <-"https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
x <- read.csv(url, row.names=1)
head(x)
```


Some base figures

```{r}
barplot(as.matrix(x), beside=FALSE, col=rainbow(nrow(x)))

```

There is one plot that can be useful for small datasets:

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

> Main point: It can be difficult to spot major trends and patterns even in relatively small multivariate datasets (here we only have 17 dimensions, typically we have 1000s).

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`

I will take the transpose of food data so the "foods" are in the columns:

```{r}
pca <- prcomp(t(x))
summary(pca)
```

```{r}
cols <- c("orange","red","blue","darkgreen")
plot(pca$x[,1], pca$x[,2],col=cols,pch=16)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x) + aes(PC1, PC2) + geom_point(col=cols)
```

```{r}
ggplot(pca$rotation) + aes(PC1, rownames(pca$rotation)) + geom_col()
```

PCA looks super useful and we will come back to describe this further next day :)
